### 2019-02-03 ic-journal/contraction.tex
text

The constraints \hl{on equations (which must be both guarded and
sequential)} in Theorem~\ref{t:Mil89} can be
weakened if we move from equations to a special kind of inequations called
  \emph{contractions}.

Intuitively, the bisimilarity contraction $\mcontrBIS$ is a preorder in which 
$P \mcontrBIS Q  $  holds  if $P \wb Q$ and, in addition, 
$Q$ \emph{has the possibility of being at least as efficient as $P$} (as far as
$\tau$-actions \hl{are} performed). 
Process $Q$, however, may be nondeterministic and may have other ways
of doing the same work, and those could be slow (i.e., involving
more $\tau$-steps than those performed by $P$).
% Thus, in contrast with expansion,  we cannot really say that `$Q$ is more efficient than
% $P$'.

\begin{definition}%[bisimulation contraction]
\label{d:BisCon}
A process relation ${\R}$ 
 is a \textbf{(bisimulation) contraction} if, whenever
 $P\RR Q$, %for all $\mu$

\begin{enumerate}
\item   $P \arr\mu P'$ implies there is $Q'$ such that $Q \arcap \mu
  Q'$ and $P' \RR Q'$;
\item $Q \arr\mu Q'$   implies there is $P'$ such that $P \Arcap \mu
 P'$ and $P' \wb Q'$\enspace.
\end{enumerate}
\emph{Bisimilarity contraction}, written as $P \mcontrBIS Q$ ($P$
\textbf{contracts to} $Q$), if $P\ \R\ Q$ for some contraction $\R$.
\end{definition}
% We say that $P$  {\emph{bisimilarly  contracts} $Q$, written
% $Q \mcontrBIS Q$.
% We write $\mexpaBIS$ for the inverse of $\mcontrBIS$.

@@

Bisimilarity  contraction is coarser than 
 the \emph{expansion relation} 
$\expa$ \cite{arun1992efficiency,sangiorgi2015equations}.
This is a
preorder widely used in proof techniques for bisimilarity and that 
intuitively refines bisimilarity by 
 formalising the idea of \emph{efficiency} between processes.
Clause (1) is the same in the two
preorders. But in clause (2) expansion uses 
$P \Arr \mu P'$, rather than $P \Arcap \mu P'$; 
 moreover with
contraction the final derivatives are simply required to be bisimilar.
An expansion 
$P \expa Q$
tells us  that $Q$ is always at least as efficient as $P$, whereas  the
 contraction $P \mcontrBIS Q$  just says that $Q$ has the  possibility of
being at least as efficient as $P$.

@@

@@

Like (weak) bisimilarity and expansion, contraction is
preserved by all operators but (direct) sum.

% next file: unique.tex

>>>>

The constraints on the unique-solution Theorem~\ref{t:Mil89} can be 
weakened if we move from equations to a special kind of inequations called
  \emph{contractions}.

Intuitively, the bisimilarity contraction $\mcontrBIS$ is a preorder in which 
$P \mcontrBIS Q  $  holds  if $P \wb Q$ and, in addition, 
$Q$ has the \emph{possibility} of being at least as efficient as $P$ (as far as
$\tau$-actions performed). 
Process $Q$, however, may be nondeterministic and may have other ways
of doing the same work, and these could be  slow (i.e., involving
more $\tau$-steps than those performed by $P$).
% Thus, in contrast with expansion,  we cannot really say that `$Q$ is more efficient than
% $P$'.

\begin{definition}%[bisimulation contraction]
\label{d:BisCon}
A process relation ${\R}$ 
 is a \emph{(bisimulation) contraction} if, whenever
 $P\RR Q$, %for all $\mu$

\begin{enumerate}
\item   $P \arr\mu P'$ implies there is $Q'$ such that $Q \arcap \mu
  Q'$ and $P' \RR Q'$;
\item $Q \arr\mu Q'$   implies there is $P'$ such that $P \Arcap \mu
 P'$ and $P' \wb Q'$\enspace.
\end{enumerate}
\emph{Bisimilarity contraction}, written as $P \mcontrBIS Q$ ($P$
\emph{contracts to} $Q$), if $P\ \R\ Q$ for some contraction $\R$.
\end{definition}
% We say that $P$  {\emph{bisimilarly  contracts} $Q$, written
% $Q \mcontrBIS Q$.
% We write $\mexpaBIS$ for the inverse of $\mcontrBIS$.

@@

Bisimilarity  contraction is coarser than 
 the \emph{expansion relation} 
$\expa$ \cite{arun1992efficiency,sangiorgi2015equations}.
This is a
preorder widely used in proof techniques for bisimilarity and that 
intuitively refines bisimilarity by 
 formalising the idea of `efficiency' between processes.
Clause (1) is the same in the two
preorders. But in clause (2) expansion uses 
$P \Arr \mu P'$, rather than $P \Arcap \mu P'$; 
 moreover with
contraction the final derivatives are simply required to be bisimilar.
An expansion 
$P \expa Q$
tells us  that $Q$ is always at least as efficient as $P$, whereas  the
 contraction $P \mcontrBIS Q$  just says that $Q$ has the  possibility of
being at least as efficient as $P$.

@@

\begin{definition}%[expansion]
\label{d:expa}
A process relation ${\R}$
  is an \emph{expansion} if, whenever
we have $P\RR Q$, for all $\mu$
 \begin{enumerate}
 \item   $P \arr\mu P'$ implies that there is $Q'$ with $Q \arcap \mu
   Q'$
  and $P' \RR Q'$;
 \item
     $Q \arr\mu Q'$   implies that there is $P'$ with $P \Arr \mu
  P'$ and $P'
 \RR Q'$.
 \end{enumerate}
  $P$  {\em expands} $Q$, written as
 $P  \expa Q$,
 if $P \RR Q$ for some expansion $\R$.
 \end{definition}

As bisimilarity, so the expansion preorder is preserved by all operators but  sums.

@@

\hl{Like} (weak) bisimilarity and expansion, contraction is
preserved by all operators but (direct) sum.

### 2019-02-03 ic-journal/equa.tex
text

\item \emph{(strongly) guarded} if, in each $E_i$, each occurrence of
  an \behav\  variable is underneath a \textbf{visible} prefix;

@@

% The proof exploits an invariance property on immediate transitions for
% guarded and sequential expressions, and then extracts a bisimulation
% (up-to bisimilarity) out
% of the solutions of the system.  
To see the  need of the sequentiality  condition, consider
 the equation (from \cite{Mil89}) $X = \res a (a. X | \outC a)$
where $X$ is guarded but not sequential. Any process that does not use $a$ is a solution.

>>>>

\item \emph{(strongly) guarded} if, in each $E_i$, each occurrence of
  an \behav\  variable is underneath a \emph{visible} prefix;

@@

% The proof exploits an invariance property on immediate transitions for
% guarded and sequential expressions, and then extracts a bisimulation
% (up-to bisimilarity) out
% of the solutions of the system.  
To see the  need of the sequentiality  condition, consider
 the equation (from \cite{Mil89}) $X = \res a (a. X | \outC a)$
where $X$ is guarded but not sequential. Any \hl{process that} does not use $a$ is a solution.

### 2019-02-03 ic-journal/expansion.tex
text

The equivalences we consider here are mainly \emph{weak} ones, in that they
abstract from the number of internal steps being performed:
\begin{definition}%[bisimilarity]
\label{d:wb}
A process relation ${\R}$ is a \textbf{bisimulation} if, whenever
 $P\RR Q$, \hl{for all $\mu\in \mathscr{L}\cup\{\tau\}$} we have:
\begin{enumerate}
\item $P \arr\mu P'$ implies that there is $Q'$ such that $Q \Arcap \mu Q'$ and $P' \RR Q'$;
\item $Q \arr\mu Q'$,implies that there is $P'$ such that $P \Arcap
  \mu P'$ and $P' \RR Q'$\enspace.
\end{enumerate}
 $P$ and $Q$ are \textbf{bisimilar},
written as $P \wb Q$, if $P \RR Q$ for some bisimulation $\R$.
\end{definition}

We sometimes call bisimilarity the \emph{weak} one, to
distinguish it from \emph{strong} bisimilarity ($\sim$),
obtained by replacing in the above definition   the weak answer $
Q\Arcap\mu Q'$ with the strong  $Q \arr \mu Q'$.
Weak bisimilarity is not preserved by the sum operator (except for
guarded sums). For this, Milner introduced \emph{observational congruence}, also called \emph{rooted
  bisimilarity} \cite{Gorrieri:2015jt,Sangiorgi:2011ut}:
\begin{definition}%[rooted bisimilarity]
\label{d:rootedBisimilarity}
Two processes $P$ and $Q$ are \textbf{rooted bisimilar}, written as $P
\rapprox Q$, iff \hl{for all $\mu\in \mathscr{L}\cup\{\tau\}$}
\begin{enumerate}
 \item  $P \arr\mu P'$ implies that there is $Q'$ such that $Q
   \Arr\mu Q'$ and $P' \wb Q'$;
 \item  $Q \arr\mu Q'$ implies that there is $P'$ such that $P
   \Arr\mu P'$ and $P' \wb Q'$\enspace.
\end{enumerate}
\end{definition}

Besides reducing the rooted bisimiarity of two processes to
the bisimilarities of their first-step transition ends, above definition also brings a proof technique for proving the rooted bisimiarity by constructing a bisimulation:
\begin{lemma}{(Rooted bisimilarity by constructing a bisimulation)}
\label{l:obsCongrByWeakBisim}
Given a (weak) bisimulation $\RR$, if two processes $P$ and $Q$
satisfies the following properties:
\begin{enumerate}
\item $P \arr\mu P'$ implies that there is $Q'$ such that $Q
   \Arr\mu Q'$ and $P' \RR Q'$;
\item the converse of (1) on the actions from $Q$.
\end{enumerate}
then $P$ and $Q$ are rooted bisimilar, i.e.~$P \approx^c Q$.
\end{lemma}

@@

\subsection{Expansions}
\label{s:expa}

\hl{The bisimulation proof method can be enhanced by means of \emph{up-to
techniques}. One of the most useful auxiliary relations in up-to
techniques is the \emph{expansion} relation} $\expa$ \cite{arun1992efficiency,sangiorgi2015equations}.
This is an asymmetric version
of $\wb$ where $P \expa Q$ means that $P \wb Q$,
but also that $Q$ achieves the same as $P$
with no more work, i.e.~with no more $\tau$ actions.
Intuitively, if $P \expa Q$, we can think of $Q$ as being
at least as fast as $P$
or, more generally, we can think that $P$ uses at least as many resources as $Q$.
\begin{definition}%[expansion]
\label{d:expa}
A process relation ${\R}$
  is an \textbf{expansion} if, whenever
we have $P\RR Q$, for all $\mu$
 \begin{enumerate}
 \item   $P \arr\mu P'$ implies that there is $Q'$ with $Q \arcap \mu
   Q'$
  and $P' \RR Q'$;
 \item
     $Q \arr\mu Q'$   implies that there is $P'$ with $P \Arr \mu
  P'$ and $P'
 \RR Q'$.
 \end{enumerate}
  $P$  {\em expands} $Q$, written as
 $P  \expa Q$,
 if $P \RR Q$ for some expansion $\R$.
 \end{definition}

Same as bisimilarity, the expansion preorder is preserved by all operators but (direct) sums.

% next file: contraction.tex

>>>>

The equivalences we consider here are mainly \emph{weak} ones, in that they
abstract from the number of internal steps being performed:
\begin{definition}%[bisimilarity]
\label{d:wb}
A process relation ${\R}$ is a \emph{bisimulation} if, whenever
 $P\RR Q$, \DS{for all $\mu$} we have:
\begin{enumerate}
\item $P \arr\mu P'$ implies that there is $Q'$ such that $Q \Arcap \mu Q'$ and $P' \RR Q'$;
\item $Q \arr\mu Q'$,implies that there is $P'$ such that $P \Arcap
  \mu P'$ and $P' \RR Q'$\enspace.
\end{enumerate}
 $P$ and $Q$ are \emph{bisimilar},
written as $P \wb Q$, if $P \RR Q$ for some bisimulation $\R$.
\end{definition}

We sometimes call bisimilarity the \emph{weak} one, to
distinguish it from \emph{strong} bisimilarity ($\sim$),
obtained by replacing in the above definition   the weak answer $
Q\Arcap\mu Q'$ with the strong  $Q \arr \mu Q'$.
Weak bisimilarity is not preserved by the sum operator (except for
guarded sums). For this, Milner introduced \emph{observational congruence}, also called \emph{rooted
  bisimilarity} \cite{Gorrieri:2015jt,Sangiorgi:2011ut}:
\begin{definition}%[rooted bisimilarity]
\label{d:rootedBisimilarity}
Two processes $P$ and $Q$ are \emph{rooted bisimilar}, written as $P
\rapprox Q$, if
\DS{for all $\mu$}:
%  \hl{for all $\mu\in \mathscr{L}\cup\{\tau\}$}
\begin{enumerate}
 \item  $P \arr\mu P'$ implies that there is $Q'$ such that $Q
   \Arr\mu Q'$ and $P' \wb Q'$;
 \item  $Q \arr\mu Q'$ implies that there is $P'$ such that $P
   \Arr\mu P'$ and $P' \wb Q'$\enspace.
\end{enumerate}
\end{definition}

Besides reducing the rooted bisimiarity of two processes to
the bisimilarities of their  \DS{immediate derivatives}, the
above definition also brings up a proof technique \DS{for proving  rooted
bisimiarity results  by constructing a bisimulation.}
\begin{lemma}%{(Rooted bisimilarity by constructing a bisimulation)}
\label{l:obsCongrByWeakBisim}
Given a (weak) bisimulation $\RR$, if two processes $P$ and $Q$
satisfies the following properties:
\begin{enumerate}
\item $P \arr\mu P'$ implies that there is $Q'$ such that $Q
   \Arr\mu Q'$ and $P' \RR Q'$;
\item the converse of (1) on the actions from $Q$.
\end{enumerate}
then $P$ and $Q$ are rooted bisimilar, i.e.~$P \approx^c Q$.
\end{lemma}

@@

### 2019-02-03 ic-journal/formal.tex
text

\textbf{In this
formalisation we consider only single-variable equations/contractions.}
This considerably
 simplifies the required proofs in HOL, also enhances the readability of
 proof scripts without loss of generality. For the informal pencil-and-paper proofs, the
 multi-variable case is just a routine adaptation, \hl{while the elegant formalisation
   of multi-variable equations is quite a challenge.} In Section~\ref{s:concl} some ideas
 are discussed.

>>>>

\emph{In this
formalisation we consider only single-variable equations/contractions.}
This considerably
 simplifies the required proofs in HOL, and  enhances the readability of
 proof scripts without loss of generality. For the  pencil-and-paper proofs, the
 multi-variable case is just a routine adaptation, \DS{while the   formalisation
 is not (this is also discussed  in Section~\ref{s:concl}).}

### 2019-02-03 ic-journal/intro.tex
text

\hl{In Milner's versions of unique-solution theorems} for proving that all
solutions are weakly (or rooted) bisimilar (in practice these are the most
relevant cases), however, the proof method has severe syntactical limitations:
the equations must be \emph{(strongly) guarded and sequential}. That is, the
variables of the equations may only be used underneath a visible
prefix and proceed, in the syntax tree, only by the sum and prefix operators.
One way of overcoming such limitations is to replace the equations
 with special inequations called
\emph{contractions} \cite{sangiorgi2015equations,sangiorgi2017equations}. Contraction is a
preorder that, roughly, places some efficiency
constraints on processes.  The uniqueness of solutions of a system of contractions
is defined as with systems of equations: any two solutions must be \hl{(componentwise)} bisimilar.
The difference with equations is in the meaning of a solution:
in the case of contractions the solution is evaluated with respect to
the contraction preorder, rather than bisimilarity. 
With contractions, most syntactic limitations of the unique-solution theorem can be
\hl{eliminated}.  One constraint that still remains in
\cite{sangiorgi2017equations} (in which the issue is bypassed using a more
restrictive CCS syntax)
is the occurrences of direct sums, due to the failure of the
substitutivity of contraction under direct sums.

The main goal of the work described in this paper is a rather
comprehensive formalisation of the core theory of CCS in the HOL
theorem prover (HOL4),  with a focus on the theory of unique solutions
of \hl{equations and} contractions.
The formalisation, however, is not confined to the theory of unique
solutions of equations \hl{(and contractions)}, but embraces a
significant portion the theory of CCS \cite{Mil89}, mostly because the
theory of unique solutions relies on a large number of more fundamental results.
Indeed the formalisation encompasses the basic properties of strong and weak
bisimilarity (e.g.~the fixed-point and substitutivity properties), the
basic properties of
rooted bisimilarity (the congruence induced by weak
bisimilarity, also called observation congruence), and
their algebraic laws. Further extensions (beyond Nesi
\cite{Nesi:1992ve}) include four versions of ``bisimulation up to''
techniques (e.g.~bisimulation up-to bisimilarity) \cite{Mil89,sangiorgi1992problem}, and the
expansion and contraction preorder (two
efficiency-like refinements of weak bisimilarity). Concerning rooted bisimilarity, the formalisation
includes Hennessy Lemma and Deng Lemma (Lemma 4.1 and 4.2 of
\cite{Gorrieri:2015jt}),
 and two long proofs saying \hl{that} the rooted bisimilarity is the coarsest (largest)
 congruence contained in (weak) bisimilarity: one following Milner's
 book \cite{Mil89}, with the hypothesis that no processes can use up
 all labels;
the other without such hypothesis, essentially formalising van Glabbeek's paper \cite{van2005characterisation}.
Similar theorems are proved for the rooted contraction preorder.
In this respect, the work is an extensive experiment with the use of the HOL theorem prover with its
recent developments, including a package for coinductive definitions.

>>>>

\DS{
In Milner's versions of the unique-solution theorems for weak
equivalences,
 the  method has severe syntactical limitations:}
the equations must be \emph{(strongly) guarded and sequential}. That is, the
variables of the equations may only be used underneath a visible
prefix and proceed, in the syntax tree, only by the sum and prefix operators.
One way of overcoming such limitations is to replace the equations
 with special inequations called
\emph{contractions} \cite{sangiorgi2015equations,sangiorgi2017equations}. Contraction is a
preorder that, roughly, places some efficiency
constraints on processes.  The uniqueness of solutions of a system of contractions
is defined as with systems of equations: any two solutions must be \hl{(componentwise)} bisimilar.
The difference with equations is in the meaning of a solution:
in the case of contractions the solution is evaluated with respect to
the contraction preorder, rather than bisimilarity. 
With contractions, most syntactic limitations of the unique-solution theorem can be
removed.  One constraint that still remains in
\cite{sangiorgi2017equations} (in which the issue is bypassed using a more
restrictive CCS syntax)
is the occurrences of direct sums, due to the failure of the
substitutivity of contraction under direct sums.

The main goal of the work described in this paper is a rather
comprehensive formalisation of the core theory of CCS in the HOL
theorem prover (HOL4),  with a focus on the theory of unique solutions
of \hl{equations and} contractions.
The formalisation, however, is not confined to the theory of unique
\DS{solutions}, but embraces a
significant portion the theory of CCS \cite{Mil89}, mostly because the
theory of unique solutions relies on a large number of more fundamental results.
Indeed the formalisation encompasses the basic properties of strong and weak
bisimilarity (e.g.~the fixed-point and substitutivity properties), the
basic properties of
rooted bisimilarity (the congruence induced by weak
bisimilarity, also called observation congruence), and
their algebraic laws. Further extensions (beyond Nesi
\cite{Nesi:1992ve}) include four versions of ``bisimulation up to''
techniques (e.g.~bisimulation up-to bisimilarity) \cite{Mil89,sangiorgi1992problem}, and the
expansion and contraction preorder (two
efficiency-like refinements of weak bisimilarity). Concerning rooted bisimilarity, the formalisation
includes Hennessy Lemma and Deng Lemma (Lemma 4.1 and 4.2 of
\cite{Gorrieri:2015jt}),
 and two long proofs saying \hl{that}  rooted bisimilarity is the coarsest (largest)
 congruence contained in (weak) bisimilarity: one following Milner's
 book \cite{Mil89}, with the hypothesis that no processes can use up
 all labels;
the other without such hypothesis, essentially formalising van Glabbeek's paper \cite{van2005characterisation}.
Similar theorems are proved for the rooted contraction preorder.
In this respect, the work is an extensive experiment with the use of the HOL theorem prover with its
recent developments, including a package for coinductive definitions.
